---
title: "Descriptive Analysis of a Multilevel Data Set"
author: "Katherine Wolf"
date: "February 20, 2020"
output:
    pdf_document:
    latex_engine: xelatex
mainfont: Calibri
monofont: Lucida Console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)

```


```{r my code}

# need to call the libraries every time you begin a new R session
library(dplyr)
library(ggplot2)
library(nnet)
library(tidyverse)
library(tableone)
library(xtable)
library(knitr)
library(tableone)
library(kableExtra)

# read in data – suppose the file dataset.csv contains continuous variables var1 and var2, and a binary variable outcome
nyses_raw_data <- read_csv("NYSES data for class.csv")

# make working data file
nyses_to_edit <- nyses_raw_data

# id
nyses_to_edit$id <- nyses_to_edit$QKEY2

# community district
nyses_to_edit$`Community district` <- nyses_to_edit$cd

# neighborhood median income
nyses_to_edit$`Neighborhood median income ($)` <- nyses_to_edit$medinc

# neighborhood percent below poverty
nyses_to_edit$`Neighborhood poverty (%)` <- nyses_to_edit$pbelowpv

# borough
borough_labels <- 
  c("Bronx",
    "Brooklyn", 
    "Manhattan", 
    "Queens", 
    "Staten Island") # (value = order)

nyses_to_edit$Borough <- 
  factor(nyses_to_edit$boro, 
         labels = borough_labels)

# age
age_labels <- 
  c("18-24",
    "25-34", 
    "35-44", 
    "45-54", 
    "55-64", 
    "65+") # (value = order)

nyses_to_edit$`Age (years)` <- 
  factor(nyses_to_edit$agecat, 
         labels = age_labels)

# race/ethnicity
race_labels <- 
  c("White",
    "African American", 
    "Asian", 
    "Hispanic/Latinx", 
    "Other") # (value = order)

nyses_to_edit$`Race/ethnicity` <- 
  factor(nyses_to_edit$racecat, labels = race_labels)

# education
ed_labels <- 
  c("Less than high school",
    "High school/GED", 
    "Some college", 
    "College graduate", 
    "Graduate work") # (value = order)

nyses_to_edit$`Education` <- 
  factor(nyses_to_edit$edcat, labels = ed_labels)

# income
income_labels <- 
  c("<= 40,000",
    "40,001 to 80,000", 
    "> 80,000") # (value = order)

nyses_to_edit$`Income ($)` <- 
  factor(nyses_to_edit$inc3cat, labels = income_labels)

# binge drinking
binge_labels <- 
  c("No", 
    "Yes") # (value = order)

nyses_to_edit$`Binge drinking` <- 
  factor(nyses_to_edit$binge, 
         labels = binge_labels)

nyses_analyze <- 
  nyses_to_edit %>% 
  select(id, 
         `Community district`,
         `Neighborhood median income ($)`, 
         `Neighborhood poverty (%)`, 
         Borough, 
         `Age (years)`, 
         `Race/ethnicity`, 
         Education, 
         `Income ($)`, 
         `Binge drinking`)

```

# Overall report

Detailed summary statistics are in Table 1.  The univariate analyses show a dataset of 4000 study participants, of whom 40 were missing data on age, 112 were missing data on race, 77 were missing data on education, and 580 were missing data on income.

Participants appear unevenly distributed among the boroughs, with Staten Island particularly poorly represented.  Middle-aged participants between 35 and 54 appear somewhat more represented than those in other groups.  A plurality of participants are white (40%), followed by African American (26.4%) and Hispanic/Latinx (24.0%).  The most common educational level was completed high school.  A plurality of participants (40%) had the lowest income level.  (Additional analyses should break out joint covariate distributions, for example, neighborhood income and percentage in poverty by race.)

In the bivariate analyses, 89% (n = 3562) of participants did not binge drink and 11% (n = 438) did.  Neighborhood income and neighborhood percent in poverty look similar between the two groups, although they are highly correlated (Pearson correlation coefficient -0.89), indicating that I should not use them both in one model.  Binge drinking appeared particularly prevalent among participants in Manhattan compared to other boroughs and in the 25-34 year-old age group (subsequently declining with age), among college graduates, and at higher income levels.


\pagebreak


```{r make a table one}

# create a list of variables for the table 
# (not including the stratification variable)
table_one_variables <- c("Neighborhood median income ($)",
                         "Neighborhood poverty (%)",
                         "Borough", 
                         "Age (years)", 
                         "Race/ethnicity", 
                         "Education", 
                         "Income ($)")

# create a list of which ones are categorical (factor)
factor_variables <- c("Borough",
                      "Age (years)", 
                      "Race/ethnicity", 
                      "Education", 
                      "Income ($)")

table_1 <- CreateTableOne(vars = table_one_variables, 
                          factorVars = factor_variables,
                          strata = "Binge drinking",
                          data = nyses_analyze,
                          test = FALSE, 
                          includeNA = TRUE)

save(table_1, 
     file = "table_1.rdata")

# print(table.1) # Standard output

# Creates a formatted table, using kable from the knitr package
# Would want to clean this up for publication purposes:
hi <- kable(print(table_1,
                  showAllLevels = TRUE,
                  printToggle = FALSE,
                  noSpaces = TRUE,
                  catDigits = 1,
                  contDigits = 1),
            col.names = c("Level", "No binge drinking", "Binge drinking"),
            caption=paste("Descriptive statistics for participants in",
                          "NYSES cohort, stratified by binge drinking."))


# hi <- kable(print(table_1,
#                    showAllLevels = TRUE,
#                    printToggle = FALSE,
#                    noSpaces = TRUE,
#                    catDigits=1,
#                    contDigits=1),
#       caption=paste("Descriptive statistics for participants in",
#                     "NYSES cohort, stratified by binge drinking."))

hi

```


\pagebreak


# Univariate analyses

## Numeric summaries

```{r univariate general summary}

nyses_analyze %>%  
  summary()

```

\pagebreak

## Neighborhood median income

```{r univariate median income}

# neighborhood median income
nyses_analyze %>% 
  summarise(Mean = mean(`Neighborhood median income ($)`), 
            Median = median(`Neighborhood median income ($)`), 
            `Standard deviation` = sd(`Neighborhood median income ($)`), 
            Minimum = range(`Neighborhood median income ($)`)[1], 
            Maximum = range(`Neighborhood median income ($)`)[2])

# neighborhood median income
ggplot(data = nyses_analyze, 
       aes(y = `Neighborhood median income ($)`)) +
  geom_boxplot()

```

\pagebreak

## Neighborhood median poverty

```{r univariate poverty}

# neighborhood percent in poverty
nyses_analyze %>% 
  ungroup() %>% 
  summarise(Mean = mean(`Neighborhood poverty (%)`), 
            Median = median(`Neighborhood poverty (%)`), 
            `Standard deviation` = sd(`Neighborhood poverty (%)`), 
            Minimum = range(`Neighborhood poverty (%)`)[1], 
            Maximum = range(`Neighborhood poverty (%)`)[2])

# neighborhood median income
ggplot(data = nyses_analyze, 
       aes(y = `Neighborhood median income ($)`)) +
  geom_boxplot()

```

\pagebreak

## Borough

### Table 

```{r univariate borough}

borough_table <- 
  tibble(Category = c(borough_labels),
         Count = table(nyses_analyze$Borough, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$Borough, 
                                             useNA = "ifany")),3)*100)

borough_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = Borough)) +
  geom_bar()

```

\pagebreak 

## Age

### Table

```{r}

age_table <- 
  tibble(Category = c(age_labels, "NA"),
         Count = table(nyses_analyze$`Age (years)`, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$`Age (years)`, useNA = "ifany")),3)*100)

age_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = `Age (years)`)) +
  geom_bar()

```

\pagebreak 

## Race/ethnicity

### Table

```{r}

re_table <- 
  tibble(Category = c(race_labels, "NA"),
         Count = table(nyses_analyze$`Race/ethnicity`, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$`Race/ethnicity`, 
                                             useNA = "ifany")),3)*100)

re_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = `Race/ethnicity`)) +
  geom_bar()

```

\pagebreak

## Education

### Table

```{r}

ed_table <- 
  tibble(Education = c(ed_labels, "NA"),
         Count = table(nyses_analyze$Education, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$Education, 
                                             useNA = "ifany")),3)*100)

ed_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = Education)) +
  geom_bar()

```

\pagebreak

## Income

### Table

```{r}

income_table <- 
  tibble(Category = c(income_labels, "NA"),
         Count = table(nyses_analyze$`Income ($)`, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$`Income ($)`, 
                                             useNA = "ifany")),3)*100)

income_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = `Income ($)`)) +
  geom_bar()

```

\pagebreak

## Binge drinking

### Table

```{r}

binge_table <- 
  tibble(Category = c(binge_labels),
         Count = table(nyses_analyze$`Binge drinking`, useNA = "ifany"), 
         Percentage = round(prop.table(table(nyses_analyze$`Binge drinking`, 
                                             useNA = "ifany")),3)*100)

binge_table

```

### Bar plot

```{r}

nyses_analyze %>% 
  ggplot(aes(x = `Binge drinking`)) +
  geom_bar()

```

\pagebreak

# Bivariate analyses by outcome

## Neighborhood median income

### Table

```{r}

# neighborhood median income
nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  summarise(Mean = mean(`Neighborhood median income ($)`), 
            Median = median(`Neighborhood median income ($)`), 
            `Standard deviation` = sd(`Neighborhood median income ($)`), 
            Minimum = range(`Neighborhood median income ($)`)[1], 
            Maximum = range(`Neighborhood median income ($)`)[2])

```

### Box plots

```{r}

# neighborhood median income
ggplot(data = nyses_to_edit, 
       aes(x = `Binge drinking`, 
           y = `Neighborhood median income ($)`)) +
  geom_boxplot()

```

\pagebreak 

## Neighborhood percent in poverty

### Table

```{r}

# neighborhood percent in poverty
nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  summarise(Mean = mean(`Neighborhood poverty (%)`), 
            Median = median(`Neighborhood poverty (%)`), 
            `Standard deviation` = sd(`Neighborhood poverty (%)`), 
            Minimum = range(`Neighborhood poverty (%)`)[1], 
            Maximum = range(`Neighborhood poverty (%)`)[2])

```

### Box plots

```{r}

# neighborhood median income
ggplot(data = nyses_to_edit, 
       aes(x = `Binge drinking`, 
           y = `Neighborhood poverty (%)`)) +
  geom_boxplot()

```

### Correlation between neighborhood poverty and neighborhood income

```{r}

cor.test(nyses_analyze$`Neighborhood median income ($)`, 
         nyses_analyze$`Neighborhood poverty (%)`)

```

\pagebreak

## Borough

### Table

```{r}

basic_two_borough <- table(nyses_analyze$Borough, 
                           nyses_analyze$`Binge drinking`, 
                           useNA = "ifany")

prop_col_borough <- round(prop.table(basic_two_borough, 2), 3) * 100

prop_row_borough <- round(prop.table(basic_two_borough, 1), 3) * 100

full_two_table_borough <- 
  tibble(Borough = c(borough_labels),
         No = basic_two_borough[,1], 
         `% No (row)` = prop_row_borough[,1], 
         `% No (column)` = prop_col_borough[,1],
         Yes = basic_two_borough[,2], 
         `% Yes (row)` = prop_row_borough[,2], 
         `% Yes (column)` = prop_col_borough[,2])

full_two_table_borough

```

### Bar plot

```{r}

nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  ggplot(aes(x = Borough, col = `Binge drinking`)) +
  geom_bar(position = position_dodge())

```

\pagebreak

## Age

### Table

```{r}

basic_two_age <- table(nyses_analyze$`Age (years)`, 
                           nyses_analyze$`Binge drinking`, 
                           useNA = "always")

prop_col_age <- round(prop.table(basic_two_age, 2), 3) * 100

prop_row_age <- round(prop.table(basic_two_age, 1), 3) * 100

full_two_table_age <- 
  tibble(`Age (years)` = c(age_labels, NA),
         No = basic_two_age[,1], 
         `% No (row)` = prop_row_age[,1], 
         `% No (column)` = prop_col_age[,1],
         Yes = basic_two_age[,2], 
         `% Yes (row)` = prop_row_age[,2], 
         `% Yes (column)` = prop_col_age[,2])

full_two_table_age

```

### Bar plot

```{r}

nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  ggplot(aes(x = `Age (years)`, col = `Binge drinking`)) +
  geom_bar(position = position_dodge())

```

\pagebreak

## Race/ethnicity

### Table

```{r}

basic_two_race <- table(nyses_analyze$`Race/ethnicity`, 
                           nyses_analyze$`Binge drinking`, 
                           useNA = "ifany")

prop_col_race <- round(prop.table(basic_two_race, 2), 3) * 100

prop_row_race <- round(prop.table(basic_two_race, 1), 3) * 100

full_two_table_race <- 
  tibble(`Race/ethnicity` = c(race_labels, NA),
         No = basic_two_race[,1], 
         `% No (row)` = prop_row_race[,1], 
         `% No (column)` = prop_col_race[,1],
         Yes = basic_two_race[,2], 
         `% Yes (row)` = prop_row_race[,2], 
         `% Yes (column)` = prop_col_race[,2])

full_two_table_race

```

### Bar plot

```{r}

nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  ggplot(aes(x = `Race/ethnicity`, col = `Binge drinking`)) +
  geom_bar(position = position_dodge())

```

\pagebreak

## Education

### Table

```{r}

basic_two_education <- table(nyses_analyze$Education, 
                           nyses_analyze$`Binge drinking`, 
                           useNA = "ifany")

prop_col_education <- round(prop.table(basic_two_education, 2), 3) * 100

prop_row_education <- round(prop.table(basic_two_education, 1), 3) * 100

full_two_table_education <- 
  tibble(Education = c(ed_labels, NA),
         No = basic_two_education[,1], 
         `% No (row)` = prop_row_education[,1], 
         `% No (column)` = prop_col_education[,1],
         Yes = basic_two_education[,2], 
         `% Yes (row)` = prop_row_education[,2], 
         `% Yes (column)` = prop_col_education[,2])

full_two_table_education

```

### Bar plot

```{r}

nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  ggplot(aes(x = Education, col = `Binge drinking`)) +
  geom_bar(position = position_dodge())

```

\pagebreak

## Income

### Table

```{r}

basic_two_income <- table(nyses_analyze$`Income ($)`, 
                           nyses_analyze$`Binge drinking`, 
                           useNA = "ifany")

prop_col_income <- round(prop.table(basic_two_income, 2), 3) * 100

prop_row_income <- round(prop.table(basic_two_income, 1), 3) * 100

full_two_table_income <- 
  tibble(Income = c(income_labels, NA),
         No = basic_two_income[,1], 
         `% No (row)` = prop_row_income[,1], 
         `% No (column)` = prop_col_income[,1],
         Yes = basic_two_income[,2], 
         `% Yes (row)` = prop_row_income[,2], 
         `% Yes (column)` = prop_col_income[,2])

full_two_table_income

```

### Bar plot

```{r}

nyses_analyze %>% 
  group_by(`Binge drinking`) %>% 
  ggplot(aes(x = `Income ($)`, col = `Binge drinking`)) +
  geom_bar(position = position_dodge())

```


```{r ref.label=knitr::all_labels(), echo = T, eval = F}

```

